---
title: "Casual vs. Member Usage"
author: "Angelo Morelli"
format: html
execute:
  echo: false
  message: false
  warning: false
---

## Casual vs. Member Usage

This page examines the differences in casual CitiBike usage vs. how CitiBike members use their bikes. Ride durations, hourly patterns, and average ride times uncover insights on how these two groups differ in usage.

```{r}
#| label: setup

library(tidyverse)
library(ggplot2)
library(scales)
```

```{r}
#| cache: TRUE

june_data1 <- read_csv("202406-citibike-tripdata_1.csv")
june_data2 <- read_csv("202406-citibike-tripdata_2.csv")
june_data3 <- read_csv("202406-citibike-tripdata_3.csv")
june_data4 <- read_csv("202406-citibike-tripdata_4.csv")
june_data5 <- read_csv("202406-citibike-tripdata_5.csv")
june_data <- bind_rows(june_data1, june_data2, june_data3, june_data4, june_data5)

december_data1 <- read_csv("202412-citibike-tripdata_1.csv")
december_data2 <- read_csv("202412-citibike-tripdata_2.csv")
december_data3 <- read_csv("202412-citibike-tripdata_3.csv")
december_data <- bind_rows(december_data1, december_data2, december_data3)

trip_data <- bind_rows(june_data, december_data)
```

# Distribution of Ride Duration

This density plot displays a main difference in how these bikes are typically used, and that would be (most likely) the commuter vs. leisure crowd. Members show a higher density in shorter rides because they are usually taking shorter trips, more focused on transport than sightseeing.

```{r}
trip_data <- trip_data |>
  mutate(
    ride_duration = as.numeric(difftime(ended_at, started_at, units = "mins")),  # Duration in minutes
    hour_of_day = as.numeric(format(started_at, "%H"))  # Hour of the day (0-23)
  )
```

```{r}
# Plot ride duration distribution
ggplot(trip_data, aes(x = ride_duration, fill = member_casual)) +
  geom_density(alpha = 0.5) +  # Density plot with transparency
  labs(title = "Distribution of Ride Durations (Members vs. Casual Users)",
       subtitle = "Area under each curve is 1",
       x = "Ride Duration (Minutes)",
       y = "Relative Frequency (Density)",
       fill = "User Type") +
  scale_x_continuous(limits = c(0, 120)) +  # Limit x-axis to 0-120 minutes for better visualization
  theme_minimal()
```

# Hourly Bike Usage

This bar chart shows the number of rides taken by hour, grouped by user type. Members exhibit a strong commuting pattern, with peaks in the early morning and late afternoon. Casual users display a more even distribution throughout the day, with higher activity during midday, though still following the general pattern.

```{r}
# Aggregate rides by hour of the day and user type
hourly_usage <- trip_data |>
  group_by(hour_of_day, member_casual) |>
  summarise(total_rides = n(), .groups = "drop")

# Plot hourly usage
ggplot(hourly_usage, aes(x = hour_of_day, y = total_rides, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +  # Bar plot with grouped bars
  labs(title = "Hourly Bike Usage (Members vs. Casual Users)",
       x = "Hour of the Day",
       y = "Total Rides",
       fill = "User Type") +
  scale_x_continuous(breaks = 0:23, labels = paste0(0:23, ":00")) +  # Format x-axis labels 
  scale_y_continuous(labels = scales::comma)
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```

# Average Ride Duration by Hour of the Day

This plot explores average ride duration for both types of riders at all times of day, further revealing the trend of longer, more lax rides for casuals and shorter, more focused rides for members.

```{r}
# Aggregate average ride duration by hour of the day and user type
avg_duration_by_hour <- trip_data |>
  group_by(hour_of_day, member_casual) |>
  summarise(avg_duration = mean(ride_duration, na.rm = TRUE), .groups = "drop")

# Plot average ride duration by hour
ggplot(avg_duration_by_hour, aes(x = hour_of_day, y = avg_duration, color = member_casual)) +
  geom_line(linewidth = 1) +  # Line plot
  geom_point(size = 2) +  # Add points for emphasis
  labs(title = "Average Ride Duration by Hour (Members vs. Casual Users)",
       subtitle = "Casual users ride longer than members",
       x = "Hour of the Day",
       y = "Average Ride Duration (Minutes)",
       color = "User Type") +
  scale_x_continuous(breaks = 0:23, labels = paste0(0:23, ":00")) +  # Format x-axis labels
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```

