---
title: "Monthly Demand"
author: "Angelo Morelli"
format: html
execute:
  echo: false
  message: false
  warning: false
---

## Bike Demand, by Hour of the Day & Day of the Month 

This page explores bike usage patterns between two separate months, revealing differences and trends in user behavior dependent upon time.

```{r}
#| label: setup

library(tidyverse)
library(ggplot2)
library(scales)
```

```{r}
#| cache: TRUE

june_data <- read_csv(unzip("202406-citibike-tripdata.zip"))

december_data <- read_csv(unzip("202412-citibike-tripdata.zip"))
                          
trip_data <- bind_rows(june_data, december_data)
```

# Demand by Hour of the Day

This graph compares the number of bike trips on the average day in June and December, showing how December usage is quite a bit lower on all fronts (likely due to the colder weather). They both align with the peaks of commuting hours.

```{r}
trip_data <- trip_data |>
  mutate(started_at = as.POSIXct(started_at, format = "%Y-%m-%d %H:%M:%S"))

monthly_demand <- trip_data |>
  mutate(
    month = format(started_at, "%B"),  # Extract month name (e.g., "June", "December")
    hour = as.numeric(format(started_at, "%H"))  # Extract hour as numeric (0-23)
  ) |>
  filter(month %in% c("June", "December")) |>  # Keep only June and December
  group_by(month, hour) |>
  summarise(total_trips = n(), .groups = "drop")
```

```{r}
ggplot(monthly_demand, aes(x = hour, y = total_trips, color = month)) +
  geom_line(size = 1) +  # Add line for trends
  labs(title = "Bike Demand by Hour of the Day (June vs. December)",
       x = "Hour of the Day",
       y = "Total Trips",
       color = "Month") +
  scale_x_continuous(breaks = 0:23, labels = paste0(0:23, ":00")) +  # Format x-axis labels
  scale_y_continuous(labels = unit_format(unit = "k", scale = 1e-3)) +  # Format y-axis labels as "100k", "200k", etc.
  scale_color_manual(values = c("June" = "darkblue", "December" = "darkred")) +  # Custom colors for months
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```

# Demand by Day of the Month

This graph compares the amount of bike demand by day of the month, where we can observe a couple of behaviors, such as the weekly downturn visible more in June's plotting, likely due to weekday commuters not using the bikes over the weekend, as well as lowest usage being observed during the holiday time in December.

```{r}
daily_demand <- trip_data |>
  mutate(
    month = format(started_at, "%B"),  # Extract month name (e.g., "June", "December")
    day_of_month = as.numeric(format(started_at, "%d"))  # Extract day of the month (1-31)
  ) |>
  filter(month %in% c("June", "December")) |>  # Keep only June and December
  group_by(month, day_of_month) |>
  summarise(total_trips = n(), .groups = "drop")
```

```{r}
ggplot(daily_demand, aes(x = day_of_month, y = total_trips, color = month)) +
  geom_line(size = 1) +  # Add line for trends
  labs(title = "Bike Demand by Day of the Month (June vs. December)",
       x = "Day of the Month",
       y = "Total Trips",
       color = "Month") +
  scale_x_continuous(breaks = 1:31, labels = 1:31) +  # Format x-axis labels for days of the month
  scale_y_continuous(labels = unit_format(unit = "k", scale = 1e-3)) +  # Format y-axis labels as "100k", "200k", etc.
  scale_color_manual(values = c("June" = "darkblue", "December" = "darkred")) +# Custom colors for months
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```

